AA Distributed Software Development Methodology

### 2012-06-26 front-matter.tex
add recent works, results, discussions and conclusions. whole paper reviewed.

\title{%
AA Distributed Software Development Methodology
}

\author{%
Renato~Fabbri \and Ricardo~Fabbri \and Vilson Vieira \and Alexandre Negrao \and Lucas Zambianchi
}

@@

\begin{abstract}
Sweet paper about the aa!
\end{abstract}

>>>>

\title{
AA Distributed Software Development Methodology
}

\author{%
Renato~Fabbri \and Ricardo~Fabbri \and Vilson Vieira \and Alexandre Negrao \and Lucas Zambianchi
\and Marcos Mendonca \and Danilo Shiga
}

@@

\begin{abstract}
We present a new self-regulating methodology for coordinating
distributed team work called Algorithmic Auto-regulation (AA), based
on recent social networking concepts and individual merit. Team
members take on an egalitarian role, and stay voluntarily logged into
so-called AA sessions for part of their time (e.g.\ 2 hours per day),
during which they create periodical logs --- short text sentences ---
they wish to share about their activity with the team. These logs are
publicly aggregated in a Website and are peer-validated, as in code
review. A video is generally recorded by the members to make their
work logs more understandable. This methodology is well-suited for
increasing the efficiency of distributed teams working on what is
called Global Software Development (GSD) through asynchronous
on-demand communication, reducing the need for central management,
unproductive meetings or time-consuming reports. The AA methodology
also legitimizes the activities of a distributed software team.  It
thus enables entities to have a solid means to fund these activities,
allowing for new and concrete business models to emerge for very
distributed software development. AA as a methodology is also at the
core of having self-replicating hacker initiatives. These claims are
discussed in a real case-study of running a distributed free software
hacker team called Lab Macambira.
\end{abstract}

### 2012-06-26 paper.tex
subsection-ized some sections
renamed figure cuz of apparent texlive bugs with image extensions
add recent works, results, discussions and conclusions. whole paper reviewed.

%
% Final Checklist TODO XXX 
% - Spellcheck
% - check xxx's todo's etc
%
%\input{preamble}
\usepackage{url}
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{color}
\usepackage{verbatim}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}

@@

One of the defining features of modern times is the widening geographical
distribution of software teams~\cite{}. One example is the free software
movement. \todo{citar alguns exemplos}.

However, we have noticed how difficult it is to coordinate and fund free
software on a larger scale than currently available, when teams are very
heterogeneous containing not only volutaries and very experienced developers,
but also contractors from different backgrounds and cultures.

Another problem faced by modern software companies and other collectives are frequent
ineffective meetings, which are seldom focused to the interest to any
attendant. The result is that it has become the norm to participate in too many
meetings with the laptop open, which is very unproductive at the very least.
Coders like to code, to be productive, to have their hands-on their project to
do what they're best at. They hate to have to stop for too many meetings,
and they hate to have to write lengthy reports to justify their funding.
\todo{ler mais. cacm, etc}

AA is a methodology and an associated software system for coordinating
distributed teamwork.  Team members take on an egalitarian role, and stay
voluntarily logged in the system for part of their time (e.g., 2h/day), during
which they microblog periodical "tweets" about their activity. These microblog
logs are publicly aggregated and are validated by their own peers. Through AA,
we have a methodology and an associated system to validate and enable the
activities of a distributed software team. It implicitly legitimizes financing
the expansions of the team's activity. The AA methodology is specially useful
for coordinating distributed and decentralized teamwork, providing effective
means to asyncrhonously update different team members without the need for
syncrhonous unproductive meetings.

Figure~\ref{fig:mm} summarizes our methodology.

@@

\todo{alerts work for greater consciousness of time}

\section{Related Work}
\todo{survey other methodologies such as agile etc}

\section{The AA Methodology}

\section{The AA Methodology}

\section{Conclusion}

%\section*{Acknowledgments}
%The authors would like to thank NSF and CNPq.

%{\small
%\input{paper-draft.bbl}
%%%%\bibliographystyle{splncs}
% this is a good style for drafting
%\bibliographystyle{abstract}
% this is a good style for finals
\bibliographystyle{acm}
\bibliography{personal}
%}

>>>>

%
% Final Checklist TODO XXX 
% - Spellcheck
% - check xxx's todo's etc
%
%\input{preamble}
\usepackage{url}
\usepackage{graphicx}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{color}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}

@@

One of the defining features of modern times is the widening
geographical distribution of software teams~\cite{last2003} creating
what is called Global Software Development (henceforth
GSD)~\cite{german2003}. An example is the free software
movement. Projects and institutions like Mozilla Foundation has many
employees and thousands of voluntary developers distributed around
many countries. The same is true for GNOME~\cite{german2003}, OpenBSD,
MySQL or Apache Software Foundation, to enumerate just the most active
projects\footnote{Ohloh, the open source network, have a more complete
  and constantly updated list of the most active projects on-line at
  \url{http://www.ohloh.net/}}. Along the free and open software
projects, GSD is growing popularity in every niche of software
industry as a whole, even on those distributing their software with
proprietary licenses. This phenomenon is attributed to various factors
like larger labor pool, natural globalization of software companies
and foundations or even the premise of cheaper cost of
production~\cite{komi2005}.

However, dispite the advantages of GCD, we have noticed how
difficult it is to coordinate and fund free software on a larger scale
than currently available, when teams are very heterogeneous containing
not only volunteers and very experienced developers, but also
contractors from different backgrounds and cultures. Our observations
found base on the factors suggested by Carmel~\cite{carmel1999} as
main difficulties for GCD: distance, time and cultural differences. In
the case of free or open software projects, all these factors are
involved.

Another problem faced by modern software companies and other
collectives are frequent ineffective meetings, which are seldom
focused to the interest to any attendant. The result is that it has
become the norm to participate in too many meetings with the ``laptop
open'', which is unproductive at the very least. Software developers
like to code, to be productive, to have their hands on their project,
to do what they are best at. They dislike to have to stop for meetings
or to write lengthy reports to justify their funding.
%\todo{ler mais. cacm, etc}

Here we propose AA, a methodology and an associated software system
for coordinating distributed team work dealing with the disadvantages
of GSD. Team members take on an egalitarian role, and stay
voluntarily \textit{logged} in the system for part of their time
(e.g.\ 2 hours per day), during which they log a periodical short text
sentence --- similar to a ``tweet'' in this Twitter era --- as the
status of their activity using an easy to use command. These
``microblog sentences'' are publicly aggregated and validated by other
team members. Through AA, we have a methodology and an associated
system to validate and enable the activities of a distributed software
team. It implicitly legitimizes financial support for the expansion of
the activity of the developer team. The AA methodology is specially
useful for coordinating distributed and decentralized team work,
providing effective means to asynchronously update different team
members without the need for synchronous unproductive meetings.

%\todo{alerts work for greater consciousness of time}
%\todo{Integrate notes from Ricardo+Renato april 15/16 meeting}
%\todo{integrate IRC chat notes}

A brief overview of current papers about GSD methodologies related with
AA is presented in Section~\ref{related-work}, while in
Section~\ref{aa-methodology} we describe the most relevant
characteristics of the AA methodology followed by the description of
our experience using AA in a team of 9 developers since July of
2011. Finally, in Section~\ref{conclusions}, we draw some final
conclusions and indicate future possibilities to the
practical use of AA on other teams of software developers or
individuals working on non-software activities.

\section{Related Work}
\label{related-work}

%% aqui tomei como ponto de vista as metodologias associadas a GSD
%% (Global Software Development) que considero foco do AA, e sua
%% principal vantagem

%\todo{survey other methodologies such as agile etc}
%\todo{probably a good source http://agilemanifesto.org/}

There has been a large amount of research done in the area of
methodologies to deal with distributed teams of developers. We are
focusing in GSD here, however some principles involved on those
methodologies could be used on smaller teams of developers working in
the same place, time and with minor cultural differences. Moreover we
generally think on ``distributed development'' being global which is
not totally true. We even applied AA to a team that work at the same
city but on different times (more details on
Section~\ref{results}). Even smaller groups of developers working on
the same building could use GSD methodologies. A thorough survey of
these methodologies is beyond the scope of this paper. Here we present
a brief overview.

Various methodologies for GSD were built around the factors that
affect distributed team works, proposed by Carmel~\cite{carmel1999}
and comprising three distances: geographical, cultural and
temporal. Geographical distance prejudice \emph{coordination}, being
the act of integrating all the tasks distributed between
units~\cite{carmel2001}; \emph{control}, or the process to maintain
specific goals, policies or quality levels; and
\emph{communication}. All these factors are correlated, for example, a
team need to have clearly communication to work on tasks of a specific
problem. Cultural distance encompass differences on organizational and
natural culture. Spoken language, unit and ethnic values are common
forms of this distance. Some companies prefer to situate development
units in foreign locations with minimal cultural distance (e.g.\ an
American firm prefer Ireland, because of spoken language
similarity~\cite{carmel2001}). And finally the temporal distance that
prejudices synchronous communications like telephone or
videoconferences. Units of developers working on different time-zones
are concerned with managing of their agenda guided by this temporal
distance.

Targeting geographical distance, Carmel~\cite{carmel2001} suggests a
tactic to reduce intensive collaboration. His approach divides the
whole software life-cycle on levels of complexity. Each level has a
degree of collaboration. For example, some developers working on a
project with high collaboration level should use the follow-the-sun
approach: when concluding the work day, they pass their work to the
team working on another time-zone. Other tactics are suggested by the
same author to deal with the three distances, like separating foreign
units of developers in time-zone bands.

Battin et al.~\cite{battin2001} propose and argues about their
experiments using specific methodologies created for the distributed
development centers of the Motorola Company (which has over 25
software development centers worldwide). These methodologies included
constant communication with critical units, incremental integration
and scheduled based on time-zones distributed to developers on 6
countries from 3 continents.

While considering free software projects instead of companies, the
same factors are present and some methodologies
arises. German~\cite{german2003} gives a concise review of
methodologies used by the GNOME project, one of the most active free
software projects and used by companies like Sun Microsystems. It is
interesting to note a difference on viewpoint present on his paper:
German focuses the methodology description on code. He start explaining
that GNOME is separated into modules (76 on version 2.4, to be
precise) and each module has one maintainer who divide his modules
into separated parts in which other developers can work on independent
tasks, along other responsibilities. As like others free software
projects, all the development was done using a \emph{bugtracker} to
bugs and issues management, mail lists and Internet Relay Chat (IRC)
to discussion and communication and a software configuration
management like SVN or Git. Periodical (commonly yearly) conferences
like GUADEC is common on free and open source projects to face-to-face
meeting and is based each time on a different place.

\section{The AA Methodology}
\label{aa-methodology}

As noted, some strategies for GSD is based on complex methodologies
and many of those were built for a specific company or software
center. Here we propose an alternative methodology based on a simple
idea: small working sessions logged by a computational
tool. Figure~\ref{fig:mm} summarizes our methodology.

@@

\subsection{AA Session}

From the developer perspective, the AA methodology is based on
creating pretty small high perspective reports of what they are doing
in a specific time frame, that can be something between 5 to 15
minutes, depending in what is more comfortable for the developer. An
AA Session would be a period of at least 2 hours doing these
reports. The developer can set reminders to show up when its time to
make a new report. The objective of the flexible time frame and
reminders is to minimize developer overhead during his AA session. In
this way he can make the reports while staying concentrated in his
code. Each report can be sent directly to an on-line server, or stored
locally in a temporarily database for sending later. This make
possible the use of AA while offline.

Developers can also record a video screencast in the end of the
session summarizing what has been done, explaining with his words and
showing his most important results. This, combined with the textual
log of his session, makes the whole report more understandable for
himself or other people searching for information about his
production.

\subsection{AA Website Report}

\begin{figure}
\begin{center}
   \includegraphics[width=0.8\linewidth,keepaspectratio=true]{figs/aa-0_1.png}
\end{center}
   \caption{AA Version 0.1}
\label{fig:aaserver}
\end{figure}

All AA reports made by the developers are sent to a Web server and
become public in a Website, being possible then for a manager or
another developer to follow very closely the work of a developer,
nearly real-time, reading each of his small reports of what he is
doing right now.

Another possibility is to check older sessions to see when sometimes
was done and the comments of the developer about it, since each AA
post happens in a very small time frame of work, the information about
what was done become very easy to understand, instead of a long
report in the end of a session.

The site is where the developer can add to his session a link for his
screencast about that session, a video summary of what was done to
complement the reports, useful specially on cases where the small
reports were done in a hurry, because the developer did not want to
lose his focus on something important at that moment.

\subsection{Validation}

Each AA session must be validated by another developer, it means that
all reports are read by someone that will consider then valid or not
and will even write commentaries about the specific session. The
developer in charge of validating a session is decided randomly by the
AA Web server, which send an email to the chosen developer with an
URL to a validation interface.

% TODO: adicionar figura aa cliente (ou tabela com os principais comandos)
% TODO: adicionar figura com arquitetura do AA cliente + AA web + validação
% TODO: melhorar screenshot do AA web

\section{Results and Discussion}
\label{results}

The easy and effective management of teams working on GSD is the main
purpose of the AA methodology. We applied this methodology to a group
of 9 developers in July of 2011 on what we called Lab
Macambira~\footnote{LabMacambira.sf.net:
  \url{http://labmacambira.sf.net}}. The main objective of the team
was to work on different free software projects, contributing directly
to their development, sending bug corrections or proposing new
features on their source code.

All the team members had different levels of knowledge on software
development as part of large and distributed free software projects
like Scilab or Mozilla. In this way, one month of training was
conducted by two experienced developers, teaching the basics of use of
development support tools like bugtrackers, programming languages and
version control systems. After this period, a challenge was proposed
for the new developers: send a bug correction or a new feature to a
large free software project and you will work with us, you will be a
``Macambira'' developer. Table~\ref{tabela:contribuicoes} summarizes
the contributions of each ``Macambira'' to free software projects.

%% adicionar tabela de contribuições

\begin{table}
  \caption{Free and open software projects that received contributions
    from ``Macambiras''. On the first column we can see a list of
    applications of those projects. At right, the pseudonym of the
    ``Macambiras'' who sent \textit{commits} to the application. At
    ``Lab Macambira'', and at free software community in general, is a
    common practice to use pseudonym as identification.}
    \begin{tabular}{|l|l|}
        \hline
        Application           & ``Commiters''                       \\
        \hline \hline
        Mozilla Firefox       & daneoshiga, bzum                    \\
        Evince                & hick209, bzum, marcicano, mquasar   \\
        BePDF / Xpdf          & marcicano                           \\
        Ekiga                 & flecha                              \\
        Empathy               & fefo                                \\
        Lib Folks (Telepathy) & kamiarc                             \\
        Scilab                & v1z, humannoise                     \\
        VxL                   & v1z                                 \\
        ImageMagick           & v1z                                 \\
        OpenOffice            & hick209                             \\
        Puredata              & v1z, automata, greenkobold, gilson, bzum \\
        Puredata OpenCV       & v1z                                 \\
        Puredata GEM          & v1z, fefo, hick209                  \\
        Puredata PDP          & v1z, fefo, hick209                  \\
        ChucK                 & rfabbri, automata                   \\
        ChucK MiniAudicle     & rfabbri, automata                   \\
        WebRTC                & automata                            \\
        OSC-Web               & automata                            \\
        Web-PD-GUI            & automata                            \\
        Live-Processing       & automata                            \\
        ChucK-Wiimote         & automata                            \\
        Audiolet              & automata                            \\
        Extempore             & automata                            \\
        \hline
    \end{tabular}
    \label{tabela:contribuicoes}
\end{table}

In one month, each developer contributed to many very large free software
projects. Many of the developers started the training with no
knowledge of what is free software and ended that period becoming a
free software developer.

During that month, the same team developed the first version of the AA
system and used AA to manage their activities. Even while developing
the system. All the source code of AA --- the client that sends the
logs and the AA Web server --- is public available~\footnote{AA source
  code:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/aa}}
and all the AA sessions log of the whole team of ``Lab Macambira'' is
also on-line~\footnote{Logs of AA sessions:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/paainel}}.

After the training period, during more 6 months, the ``Macambiras''
worked on a large range of free software projects, distributed on work
groups --- each work group focusing on a specific theme like video,
audio and web --- and financed by contracts and support of the
``Pont\~{a}o N\'{o}s Digitais''. In Table~\ref{tabela:criados} we can
see a list of the free software created by ``Lab Macambira'' since
July of 2011.

\begin{table}
    \caption{Software projects created by ``Lab Macambira'' since July
      of 2011 with a short description and the technologies --- like
      programming languages or frameworks --- involved. It is
      interesting to note the heterogeneity of projects and its areas
      of application.}
    \begin{tabular}{|l|p{5cm}|l|}
        \hline
        Application & Description & Technologies involved \\ 
        \hline \hline
        AA            & Algorithmic Auto-regulation      & Python, PHP \\
        \hline
        \'{A}gora Communs & System for on-line deliberations & PHP \\
        \hline
        SIP           & Scilab Image Processing toolbox & C, Scilab \\
        \hline
        animal        & An Imaging Library              & C \\
        \hline
        TeDi          & Test Framework for Distance Transform
        Algorithms & C, Shellscript, Scilab \\
        \hline
        Macambot      & Multi-use IRC Bot               & Python \\
        \hline
        ``Confer\^{e}ncia Permanente'' & Platform for the permanent
        conference of the rights of minors & PHP, JavaScript \\
        \hline
        CPC           & Center for accounting of the Brazilian culture
        representation groups & Python, Django \\
        \hline
        Timeline      & Interactive time lines on the Web & JavaScript
        \\
        \hline
        Imagemap      & Interactive marking for on-line photos &
        JavaScript \\
        \hline
        ABT           & Program for real-time execution and musical
        rhythmic analysis & Python \\
        \hline
        EKP           & Emotional Kernel Panic & Python, ChucK \\
        \hline
        SOS           & Aggregation and diffusion of popular and native
        knowledge about health & Python, Django \\
        \hline
        Creative Economy & Platform for creative, collaborative and
        solidarity economy of the culture hubs and cultural entities &
        Python, Django \\
        \hline
        OpenID Integration & Adaptations to existing software for
        unified login through OpenID & PHP \\
        \hline
        pAAinel & Dashboard for the real-time visualization of Lab
        Macambira activity & Python, Django \\
        \hline
        Georef & Collection of scripts to be used as reference, which
        aims to be a GIS platform to map public data of use to
        citizens & Python, Django \\
        \hline
        AirHackTable & Software for an instrument which generates
        sound from flying origami tracked by webcams & Puredata,
        C/C++, Scilab \\
        \hline
        \end{tabular}
    \label{tabela:criados}
\end{table}

As of this writing the ``Lab Macambira'' have many software
developers, and some of the trained developers continue to work
voluntarily in the project.

\section{Conclusions}
\label{conclusions}

%% brief introduction
In a scenario where GCD is growing as a popular form of software
development, not just on free software projects but in the whole
software industry, we need methodologies to deal with its
disadvantages and at the same time to amplify its advantages.

This paper has presented a methodology to GCD, being the development
conducted on large or small groups of software developers, working on
different countries or even at the same room. The AA methodology
implements a simple system where each developer take notes of his work
generating a periodical log of small text sentences. The sum of those
sentences, along an entire session of work, results in a complete
report. The report is made public available through a Website and be
validated by other developers sorted aleatory by the AA Web server.

Instead of a merely work-management tool, AA act as a methodology to
improve the time sense of individuals, dividing their work on small
sessions, and also reducing the need of extensive
reports or unnecessary meetings. By asking users to write a minimal
text sentence as a continuously log, AA does not disturb developers
concentrated on programming: developers just have to type some
characters, hit \textit{enter} and go back to coding.

AA application is not restricted to software development. As of this
writing there is a comic book studio~\footnote{Pula pirata:
  \url{http://pulapirata.com}} starting to use AA to manage their
activities. People with non-software background, like social
scientists, musicians and activists has also using AA and contributing
for its improvement.

For developer teams, we have experienced the use of AA to
auto-regulate the work of ``Lab Macambira'', a group of free software
developers from Brazil. Since July of 2011 the group have contributed
and created new free and open source software for a vast number of
applications.

There are many aspects of the work which remain unfinished. New ways
to report logs --- the ``Twitter like'' messages --- from different
interfaces like IRC, Internet Messaging services and email can make
the use of AA easy and widespread, turning AA an ubiquitous system,
presented on everyday communication channels. Even the work logs
generated since July of 2011 could be vastly statistically analyzed
aiming to recognize patterns in the behavior of individuals and their
productions.

We would like to conclude setting an important role of AA: being a
free software system and an open methodology, AA could be used to
auto-manage groups of individuals working on software or other kinds
of activities. In this way, we are interested to spread AA for those
groups, to have even more developers contributing in a collaborative
way.

\section*{Acknowledgments}
%The authors would like to thank NSF and CNPq.

We would like to also thanks AA: this whole research and even this
paper was written using AA. The complete log is on-line at
\url{http://hera.ethymos.com.br:1080/paainel/casca/}.

%{\small
%\input{paper-draft.bbl}
%%%%\bibliographystyle{splncs}
% this is a good style for drafting
%\bibliographystyle{abstract}
% this is a good style for finals
\bibliographystyle{acm}
\bibliography{aa.bib}
%}
